Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-27 paper.qmd
tidied up intro code

<!-- UMAP layout with author's suggested parameter choice-->
```{r}
#| warning: false
#| echo: false

## Import data
training_data_pbmc <- read_rds("data/pbmc3k/pbmc_pca_50.rds")
training_data_pbmc <- training_data_pbmc[, 1:9] |>
  mutate(ID = 1:NROW(training_data_pbmc))

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")
umap_pbmc_scaled <- as.data.frame(
  do.call(cbind, 
          gen_scaled_data(data = umap_pbmc, 
                          x = "UMAP1", 
                          y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'a', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3) 
```

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_1 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'b', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

@@

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.99.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_2 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'c', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.03.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_3 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'd', x = Inf, y = Inf, hjust = 1.5, vjust = 1.3, size = 3)

```{r}
#| warning: false
#| echo: false

umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_84_min_dist_0.99.rds")
umap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = umap_pbmc, x = "UMAP1", y = "UMAP2"))) |>
  dplyr::rename(c("UMAP1" = "scaled_UMAP1", 
                  "UMAP2" = "scaled_UMAP2")) |>
  dplyr::mutate(ID = 1:NROW(umap_pbmc))

umap_plot_pbmc_4 <- umap_pbmc_scaled |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'e', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc_1 <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'f', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)   
```

```{r}
#| warning: false
#| echo: false
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'g', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.1, size = 3)   
```

```{r}
#| warning: false
#| echo: false
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_92.rds")
tsne_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = tsne_pbmc, 
                                                                 x = "tSNE1", y = "tSNE2"))) |>
  dplyr::rename(c("tSNE1" = "scaled_tSNE1", 
                  "tSNE2" = "scaled_tSNE2")) |>
  dplyr::mutate(ID = 1:NROW(tsne_pbmc))

tsne_plot_pbmc_2 <- tsne_pbmc_scaled |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'h', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)   
```

```{r}
#| warning: false
#| echo: false

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_2.rds")
phate_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = phate_pbmc, x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1", 
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(phate_pbmc))

phate_plot_pbmc_1 <- phate_pbmc_scaled |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'i', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
phate_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = phate_pbmc, x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1", 
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(phate_pbmc))

phate_plot_pbmc_2 <- phate_pbmc_scaled |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'j', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_10.rds")
phate_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = phate_pbmc, x = "PHATE1", y = "PHATE2"))) |>
  dplyr::rename(c("PHATE1" = "scaled_PHATE1", 
                  "PHATE2" = "scaled_PHATE2")) |>
  dplyr::mutate(ID = 1:NROW(phate_pbmc))

phate_plot_pbmc_3 <- phate_pbmc_scaled |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'k', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
trimap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = trimap_pbmc, x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1",
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(trimap_pbmc))

trimap_plot_pbmc_1 <- trimap_pbmc_scaled |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'l', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_30_4_10.rds")
trimap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = trimap_pbmc, x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1",
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(trimap_pbmc))

trimap_plot_pbmc_2 <- trimap_pbmc_scaled |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'm', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_5_2_2.rds")
trimap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = trimap_pbmc, x = "TriMAP1", y = "TriMAP2"))) |>
  dplyr::rename(c("TriMAP1" = "scaled_TriMAP1",
                  "TriMAP2" = "scaled_TriMAP2")) |>
  dplyr::mutate(ID = 1:NROW(trimap_pbmc))

trimap_plot_pbmc_3 <- trimap_pbmc_scaled |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'n', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_10_pca_0.5_2.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_1 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'p', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_2 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'q', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_5_pca_0.1_1.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_3 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 'r', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

```{r}
#| warning: false
#| echo: false

pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.1_1.rds")
pacmap_pbmc_scaled <- as.data.frame(do.call(cbind, gen_scaled_data(data = pacmap_pbmc, x = "PaCMAP1", y = "PaCMAP2"))) |>
  dplyr::rename(c("PaCMAP1" = "scaled_PaCMAP1",
                  "PaCMAP2" = "scaled_PaCMAP2")) |>
  dplyr::mutate(ID = 1:NROW(pacmap_pbmc))

pacmap_plot_pbmc_4 <- pacmap_pbmc_scaled |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  #coord_equal() +
  annotate(geom = 'text', label = 's', x = Inf, y = Inf, hjust = 1.5, vjust = 1.5, size = 3)

```{r}
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Which is the most accurate representation of the structure in high dimensions?"
#| label: fig-diffnldrpbmc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
# 2D layouts from different NLDR techniques and different hyperparameter choices applied for the PBMC3k dataset: (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (d) UMAP (n_neighbors = 15, min_dist = 0.03), (e) UMAP (n_neighbors = 84, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (h) tSNE (perplexity = 92), (i) PHATE (knn = 2), (j) PHATE (knn = 5), (k) PHATE (knn = 10), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (m) TriMAP (n_inliers = 30, n_outliers = 4, n_random = 10), (n) TriMAP (n_inliers = 5, n_outliers = 2, n_random = 2), (p) PaCMAP (n_neighbors = 10, init = pca, MN_ratio = 0.5, FP_ratio = 2), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5), (r) PaCMAP (n_neighbors = 5, init = pca, MN_ratio = 0.1, FP_ratio = 1), and (s) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.1, FP_ratio = 1). Is there a best representation of the original data or are they all providing  equivalent information?
umap_plot_pbmc + umap_plot_pbmc_1 + umap_plot_pbmc_2 + #umap_plot_pbmc_3 + umap_plot_pbmc_4 + 
  tsne_plot_pbmc_1 + tsne_plot_pbmc + 
  #tsne_plot_pbmc_2 + phate_plot_pbmc_1 + 
  phate_plot_pbmc_2 + #phate_plot_pbmc_3 + 
  trimap_plot_pbmc_1 + #trimap_plot_pbmc_2 + trimap_plot_pbmc_3 + pacmap_plot_pbmc_1 + pacmap_plot_pbmc_2 + 
  pacmap_plot_pbmc_3 + #pacmap_plot_pbmc_4 + 
  plot_layout(ncol = 4)
```

>>>>

```{r}
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_30_min_dist_0.3.rds")

nldr1 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2)) +
  geom_point(alpha=0.1, size=1) +
  ggtitle("a")
```

```{r}
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_5_min_dist_0.01.rds")

nldr2 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("b")

@@

```{r}
umap_pbmc <- read_rds("data/pbmc3k/pbmc_umap_15_min_dist_0.99.rds")
nldr3 <- umap_pbmc |>
  ggplot(aes(x = UMAP1,
             y = UMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("c") 
```

```{r}
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_5.rds")

nldr4 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("d")  
```

```{r}
tsne_pbmc <- read_rds("data/pbmc3k/pbmc_tsne_51.rds")

nldr5 <- tsne_pbmc |>
  ggplot(aes(x = tSNE1,
             y = tSNE2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("e") 
```

```{r}
phate_pbmc <- read_rds("data/pbmc3k/pbmc_phate_5.rds")
nldr6 <- phate_pbmc |>
  ggplot(aes(x = PHATE1,
             y = PHATE2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("f")  
```

```{r}
trimap_pbmc <- read_rds("data/pbmc3k/pbmc_trimap_12_4_3.rds")
nldr7 <- trimap_pbmc |>
  ggplot(aes(x = TriMAP1,
             y = TriMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("g") 
```

```{r}
pacmap_pbmc <- read_rds("data/pbmc3k/pbmc_pacmap_30_random_0.9_5.rds")
nldr8 <- pacmap_pbmc |>
  ggplot(aes(x = PaCMAP1,
             y = PaCMAP2))+
  geom_point(alpha=0.1, size=1) +
  ggtitle("h")
```

```{r}
#| echo: false
#| fig-cap: "Six different NLDR representations of the same data. Different techniques and different parameter choices are used. Which is the most accurate representation of the structure in high dimensions?"
#| label: fig-diffnldrpbmc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
# 2D layouts from different NLDR techniques and different hyperparameter choices applied for the PBMC3k dataset: (a) UMAP (n_neighbors = 30, min_dist = 0.3), (b) UMAP (n_neighbors = 5, min_dist = 0.01), (c) UMAP (n_neighbors = 15, min_dist = 0.99), (d) UMAP (n_neighbors = 15, min_dist = 0.03), (e) UMAP (n_neighbors = 84, min_dist = 0.99), (f) tSNE (perplexity = 5), (g) tSNE (perplexity = 51), (h) tSNE (perplexity = 92), (i) PHATE (knn = 2), (j) PHATE (knn = 5), (k) PHATE (knn = 10), (l) TriMAP (n_inliers = 12, n_outliers = 4, n_random = 3), (m) TriMAP (n_inliers = 30, n_outliers = 4, n_random = 10), (n) TriMAP (n_inliers = 5, n_outliers = 2, n_random = 2), (p) PaCMAP (n_neighbors = 10, init = pca, MN_ratio = 0.5, FP_ratio = 2), (q) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.9, FP_ratio = 5), (r) PaCMAP (n_neighbors = 5, init = pca, MN_ratio = 0.1, FP_ratio = 1), and (s) PaCMAP (n_neighbors = 30, init = random, MN_ratio = 0.1, FP_ratio = 1). Is there a best representation of the original data or are they all providing  equivalent information?
nldr1 + nldr2 + nldr3 + nldr4 +
  nldr5 + nldr6 + nldr7 + nldr8 +
  plot_layout(ncol = 4)
```
